Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology

### 2023-06-18 manuscript/manuscript.qmd
float placement

All code and data for this project are available at the Open Science Framework (OSF) at https://osf.io/tg68 and permanently archived at Zenodo (doi.org/10.5281/zenodo.8052030). For our simulations and analyses we used R `r paste(R.Version()$major, R.Version()$minor, sep = ".")` [@rcoreteam2020] and we used the R package 'metafor' [@10.18637/jss.v036.i03] to perform meta-analyses and estimate heterogeneity. For parallel computing we took advantage of the R package 'parabar' [@constantin2023] and for data handling 'data.table' [@dowle2021] and 'ggplot2' [@wickham2016].

@@

@@

```{r Figure 1, out.width= "90%"}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Flowchart summarizing the simulation design. For product-moment correlations, option (1) corresponds to the Hedges and Vevea (1998) model and option (2) to the Schmidt and Hunter (2015) model (see methods section 'Meta-analytic model'). $\\theta$ = average true effect size (i.e., without sampling or measurement error), $\\tau^2$ = variance of true effect sizes, $\\theta_i$ = study true effect size, $\\gamma_i$ = study effect size with sampling error $\\sigma_i^2$, $N_i$ = study sample size, $R_i$ = study reliability, $\\bar{R}$ = average reliability across studies, $\\delta$ = standard deviation of reliabilities across studies, $r_i$ = observed study effect size with sampling and measurement error, $\\bar{r}$ = average observed effect size across studies. This figure was created using the website draw.io."

@@

```{r Figure 2, echo=FALSE, out.width = '90%'}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Unreliability in primary studies leads to a net negative bias when heterogeneity is positive. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to truncation in Pearson’s r or translation to Fisher’s z these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15, within-study sample size = 150, and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."

@@

To explore further under what conditions the forces of inflation and suppression of heterogeneity reach an equilibrium, we added an analysis for several conditions with $\tau$ < 0.1. We used the following values {0.02, 0.04, 0.06, 0.08}. Figure 3 presents the results of this analysis for Pearson's $r$. Because these lower heterogeneity levels can be expected to be affected by heterogeneity truncation to a larger extent (which results in inflation), we present results for four different numbers of studies in the meta-analysis (rows Figure 3, lower k is associated with larger inflation of estimates). The standard deviation in reliabilities is again 0.15. Figure 3 shows that even when only little heterogeneity is present, it will generally be underestimated. If μ < 0.3 then imperfect reliability leads to a negative bias in heterogeneity estimates for all $\tau$ < 0.06 in Pearson's $r$. Even for the lower levels of true heterogeneity, estimates of heterogeneity will generally be positively biased only when average effect size is large, or the number of meta-analyzed studies is small.

@@

@@

```{r Figure 4, echo=FALSE, out.width = '90%'}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "A smaller sample size in primary studies tends to exacerbate the bias in observed heterogeneity estimates due to unreliability. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Rows indicate the (fixed) sample size within studies and columns indicate nominal true between-studies standard deviation. Due to effect size truncation in Pearson's $r$ these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15 and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."

>>>>

All code and data for this project are available at the Open Science Framework (OSF) at [osf.io/tg68](https://osf.io/tg68) and permanently archived at Zenodo ([doi.org/10.5281/zenodo.8052030](https://doi.org/10.5281/zenodo.8052030)). For our simulations and analyses we used R `r paste(R.Version()$major, R.Version()$minor, sep = ".")` [@rcoreteam2020] and we used the R package 'metafor' [@10.18637/jss.v036.i03] to perform meta-analyses and estimate heterogeneity. For parallel computing we took advantage of the R package 'parabar' [@constantin2023] and for data handling 'data.table' [@dowle2021] and 'ggplot2' [@wickham2016].

@@

```{r Figure 1, out.width= "100%"}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Flowchart summarizing the simulation design. For product-moment correlations, option (1) corresponds to the Hedges and Vevea (1998) model and option (2) to the Schmidt and Hunter (2015) model (see methods section 'Meta-analytic model'). $\\theta$ = average true effect size (i.e., without sampling or measurement error), $\\tau^2$ = variance of true effect sizes, $\\theta_i$ = study true effect size, $\\gamma_i$ = study effect size with sampling error $\\sigma_i^2$, $N_i$ = study sample size, $R_i$ = study reliability, $\\bar{R}$ = average reliability across studies, $\\delta$ = standard deviation of reliabilities across studies, $r_i$ = observed study effect size with sampling and measurement error, $\\bar{r}$ = average observed effect size across studies. This figure was created using the website draw.io."

@@

@@

```{r Figure 2, echo=FALSE, out.width = '100%'}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Unreliability in primary studies leads to a net negative bias when heterogeneity is positive. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to truncation in Pearson’s r or translation to Fisher’s z these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15, within-study sample size = 150, and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."

@@

@@

To explore further under what conditions the forces of inflation and suppression of heterogeneity reach an equilibrium, we added an analysis for several conditions with $\tau$ < 0.1. We used the following values {0.02, 0.04, 0.06, 0.08}. Figure 3 presents the results of this analysis for Pearson's $r$. Because these lower heterogeneity levels can be expected to be affected by heterogeneity truncation to a larger extent (which results in inflation), we present results for four different numbers of studies in the meta-analysis (rows Figure 3, lower k is associated with larger inflation of estimates). The standard deviation in reliabilities is again 0.15. Figure 3 shows that even when only little heterogeneity is present, it will generally be underestimated. If $\mu$ < 0.3 then imperfect reliability leads to a negative bias in heterogeneity estimates for all $\tau$ < 0.06 in Pearson's $r$. Even for the lower levels of true heterogeneity, estimates of heterogeneity will generally be positively biased only when average effect size is large, or the number of meta-analyzed studies is small.

```{r Figure 4, echo=FALSE, out.width = '100%'}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "A smaller sample size in primary studies tends to exacerbate the bias in observed heterogeneity estimates due to unreliability. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations averaged across 10,000 replications. Rows indicate the (fixed) sample size within studies and columns indicate nominal true between-studies standard deviation. Due to effect size truncation in Pearson's $r$ these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line corresponds to an average reliability in primary studies. Results are for standard deviation in reliability = 0.15 and 20 studies within each meta-analysis. Code to reproduce figure: osf.io/8ygwj."

@@
